%!TeX root=../tese.tex

\chapter{Background}
\label{cap:background}

In this chapter, we present the theoretical background of DLs and some definitions.

\todo[inline]{Add initial description of the chapter}

\section{Description logics}
\todo[inline]{Add section about DLs in general}

\section{The description logic \texorpdfstring{$\elpp$}{𝓔𝓛++}}
\label{sec:el}

$\elpp$ is an extension of the DL $\el$ \citep{Baader2005a}. It was created with large bio-health ontologies in mind, such as SNOMED-CT, the NCI thesaurus, and Galen, and became an official OWL 2 profile \citep{owl2}. We concentrate on presenting $\elpp$ without concrete domains.

\subsection{Syntax}
In $\elpp$, \emph{concept descriptions} are defined inductively from a set $\Nc$ of \emph{concept names}, a set $\Nr$ of \emph{role names} and set $\Ni$ of \emph{individual names} as follows:
\begin{itemize}
	\item $\top$, $\bot$ and concept names in $\Nc$ are concept descriptions;
	\item if $C$ and $D$ are concept descriptions, $C \sqcap D$ is a concept description;
	\item if $C$ is a concept description and $r \in \Nr$, $\exists r.C$ is a concept description;
	\item if $a \in \Ni$, $\{a\}$ is a concept description.
\end{itemize}

To represent knowledge using concept descriptions, we need to define facts (axioms and role inclusions) and assertions.

An \emph{axiom}, or a \emph{general concept inclusion} (GCI), is an expression of the form $C \isa D$, where $C$ and $D$ are concept inclusions. A \emph{role inclusion} (RI) is an expression of the form $r_1 \circ \cdots \circ r_k \isa r$, where $r_1, \cdots r_k, r \in \Nr$.  The symbol ``$\circ$'' denotes composition of binary relations. A \emph{constraint box} (CBox) is a finite set of GCIs and a finite set of RIs.

Similarly, a \emph{concept assertion} is an expression of the form $C(a)$ and a \emph{role assertion}, $r(a, b)$, where $C$ is a concept description, $a, b \in \Ni$ and $r \in \Nr$. A finite set of concept assertions and role assertions is an \emph{assertional box} (ABox).

Then, an $\elpp$ \emph{knowledge base} $\K$ (KB) is a pair ($\mathcal{C}, \mathcal{A}$), where $\mathcal{C}$ is a CBox and $\mathcal{A}$ is an~ABox.

\subsection{Semantics}
The semantics of $\elpp$ are given by \emph{interpretations} $\I = (\Delta^\I, \cdot^\I)$. The \emph{domain} $\Delta^\I$ is a non-empty set of individuals, and the \emph{interpretation function} $\cdot^\I$ maps each concept name $A \in \Nc$ to a subset $A^\I$ of $\Delta^\I$, each role name $r$ to a binary relation $r^\I \subseteq \Delta^\I \times \Delta^\I$, and each individual name $a$ to an element $a^\I \in \Delta^\I$. The extension of $\cdot^\I$ for an arbitrary concept description is inductively defined by the third column of \autoref{table:elpp-syntax-semant}.

\begin{table}
	\centering
	\begin{tabular}{@{}rcc@{}}
		\toprule
		Name                    & Syntax                              & Semantics                                                                                        \\
		\midrule
		top                     & $\top$                              & $\Delta^\I$                                                                                      \\
		bottom                  & $\bot$                              & $\emptyset$                                                                                      \\
		nominal                 & $\{a\}$                             & $\{a^\I\}$                                                                                       \\
		conjunction             & $C \sqcap D$                        & $C^\I \cap D^\I$                                                                                 \\
		existential restriction & $\exists r.C$                       & $\{ x \in \Delta^\I \, | \, \exists y \in \Delta^\I \, : \, (x, y) \in r^\I \land y \in C^\I \}$ \\
		GCI                     & $C \isa D$                          & $C^\I \subseteq D^\I$                                                                            \\
		RI                      & $r_1 \circ \cdots \circ r_k \isa r$ & $r_1^\I \circ \cdots \circ r_k^\I \subseteq r^\I$                                                \\
		concept assertion       & $C(a)$                              & $a^\I \in C^\I$                                                                                  \\
		role assertion          & $r(a, b)$                           & $(a^\I, b^\I) \in r^\I$                                                                          \\
		\bottomrule
	\end{tabular}
	\caption{Syntax and semantics of $\elpp$ without concrete domains}
	\label{table:elpp-syntax-semant}
\end{table}

The interpretation $\I$ \emph{satisfies}:
\begin{itemize}
	\item an axiom $C \isa D$ if $C^\I \subseteq D^\I$ (represented as $I \models C \isa D$);

	\item a RI $r_1 \circ \cdots \circ r_k \isa r$ if $r_1^\I \circ \cdots \circ r_k^\I \subseteq r^\I$ (represented as $I \models r_1 \circ \cdots \circ r_k \isa r$);

	\item an assertion $C(a)$ if $a^\I \in C^\I$ (represented as $I \models C(a)$);

	\item an assertion $r(a, b)$ if $(a^\I, b^\I) \in r^\I$ (represented as $I \models r(a, b)$).
\end{itemize}

Also, we say that $\I$ is a \emph{model} of:
\begin{itemize}
	\item a CBox $\cbox$ if it satisfies every axiom and RI in $\cbox$ (represented as $\I \models \cbox$);
	\item an ABox $\abox$ if it satisfies every assertion in $\abox$ (represented as $\I \models \abox$);
\end{itemize}

Then, an important problem in $\elpp$ is to determine its $consistency$, that is if $\abox$ and $\cbox$ have a common model, which is in PTime \citep{Baader2005a}.

\subsection{Normal form}
We can convert an $\elpp$ knowledge base into a normal form, in polynomial time, by introducing new concept and role names \citep{Baader2005a}.

First, there is no need of explicit ABox, because $\I \models C(a)\iff \I \models \{a\} \isa C$ and $\I \models r(a, b) \iff \{a\} \isa \exists r.\{b\}$. In other words, a knowledge base can be represented by just a CBox, by transforming assertions in axioms.

In addition, given a CBox $\cbox$, consider the set $\bc$ of \emph{basic concept descriptions}, which is the smallest set of concept descriptions that contains the top concept $\top$, all concept names used in $\cbox$ and all concepts of the form $\{a\}$ used in $\cbox$.

Then, every axiom can be represented in the following normal form, where $C_1, C_2 \in \bc$, $D \in \bc \cup \{\bot\}$:
\begin{align*}
	C_1            & \isa D             \tag{simple}           \\
	C_1            & \isa \exists r.C_2 \tag{existential-head} \\
	C_2            & \isa D             \tag{existential-body} \\
	C_1 \sqcap C_2 & \isa D  \tag{conjunctive-body}
\end{align*}
And every RI are of the form $r \isa s$ or $r_1 \circ r_2 \isa s$.

\begin{example}
	\label{exmp:el-cbox-def}
	Consider the following CBox $\CC_{exa}$ representing the situation in \autoref{exmp:real-example}. On the left, we have basic knowledge of diseases and, on the right, the specific knowledge about Mary. Note that, for simplicity, it is not in normal form.

		{
			\sffamily
			\begin{center}
				\begin{minipage}{0,4\textwidth}
					\fontsize{10}{14}
					\selectfont
					Fever $\isa$ Symptom\\
					Cough $\isa$ Symptom\\
					% FeverAndCough $\isa$ Fever\\
					COVID-19 $\isa$ Disease\\
					Symptom $\isa \, \exists$hasCause.Disease\\
					Patient $\isa\, \exists$hasSymptom.Symptom\\
					CovidPatient $\equiv\, \exists$suspectOf.COVID-19\\
					hasSymptom $\circ$ hasCause $\isa$ suspectOf\\
				\end{minipage}
				\hspace{10pt}
				\begin{minipage}{0,4\textwidth}
					\fontsize{10}{14}
					\selectfont
					$\{$mary$\}$ $\isa$ Patient\\
					$\{$s1$\}$ $\isa$ Fever $\sqcap$ Cough\\
					$\{$mary$\}$ $\isa \, \exists$hasSymptom.$\{$s1$\}$\\
				\end{minipage}
			\end{center}
		}

	Because CBoxes can only represent facts, there is no way to describe uncertain knowledge. Even though, in cases when which of them are true, we could define three axioms

	\begin{description}
		\item { \sffamily $Ax_1 := $ Fever $\isa \exists$hasCause.COVID-19}, when fever is actually caused by COVID-19;
		\item { \sffamily $Ax_2 :=$ Fever $\sqcap$ Cough $\isa \exists$hasCause.COVID-19 }, when both fever and cough are caused by COVID-19;
		\item { \sffamily $Ax_3 :=$ COVID-19 $\isa \bot$}, when there are no presence of COVID-19 in the hospital.
	\end{description}

	In the following sections, it will be presented how to add these axioms in a KB with probabilistic properties.
\end{example}

We want to model uncertain information using DLs. However, it has been proved that, by adding probabilistic reasoning capabilities to $\elpp$, the complexity reaches NP-completeness \citep{Fin2019b}. Then, it is necessary to reduce the expressiveness of this language.

\section{Graphic \texorpdfstring{$\el$}{𝓔𝓛} (\texorpdfstring{$\gel$}{𝓖𝓔𝓛})}
\label{sec:gel}

\emph{Graphic} $\el$ ($\gel$) is a fragment of $\elpp$ in which every axiom and RI are in normal form and no conjunctive-body axiom is allowed \citep{Fin2020}. The semantics are the same as that of $\elpp$.

\begin{example}
	\label{exmp:gel-def}
	Since there are conjunctive-body axioms in the CBox in \autoref{exmp:el-cbox-def}, we need to modify this knowledge in order to represent it in $\gel$. First, we substitute every concept description {\sffamily Fever $\sqcap$ Cough} by a new basic concept  {\sffamily FeverAndCough}. After that, we add axioms {\sffamily FeverAndCough $\isa$ Fever} and {\sffamily FeverAndCough $\isa$ Cough} to CBox.
\end{example}

The name of this fragment comes from the fact that each GCI can be represented as arrows in a graph where nodes are basic concepts. This representation is useful for the development of algorithms and it is used to define its SAT decision.

\subsection{Graphical representation}

Consider a $\gel$-CBox $\CC$ with $n_R$ roles, its graphical representation is a edge-labeled graph $G(\CC) := (N, E, \ell)$, where $N$ is a set of nodes, $E \subseteq N^2$ is a set of directed edges and $\ell: E \to \{0, 1, \dots, n_R\}$ is a labeling function.


\todo[inline]{Do I define the graph construction here? It is the same process that I use in the implementation and I could describe it there}


\tikzstyle{block} = [draw, rectangle, rounded corners=.2cm, fill=black!5, draw=white]
\tikzstyle{dashblock} = [draw, rectangle, dashed, rounded corners=.2cm, fill=black!5, draw=black!30, line width=1pt]
\tikzstyle{arrow} = [>=latex, line width=1pt]

\begin{figure}[ht]
	\centering
	\small
	\begin{tikzpicture}[auto, node distance=1.5cm, ->]
		\tikzset{every node/.style={minimum height=0.75cm}}

		\sffamily
		\node (init) {Init};
		\node [below=of init] (s1) {$\{$s1$\}$};
		\node [below=of s1] (mary) {$\{$mary$\}$};

		\node [block, right=0.5cm of mary] (patient) {Patient};
		\node [block, right=of s1] (fevcough) {FeverAndCough};
		\node [block, below=0.5cm of fevcough] (cough) {Cough};
		\node [block, right=of fevcough] (fever) {Fever};
		\node [block, below=of fever] (symptom) {Symptom};
		\node [block, above=of fever] (covid) {COVID-19};
		\node [block, above=of covid] (covidpat) {covidPatient};
		\node [block, right=of fever] (disease) {Disease};
		\node [dashblock, left=of covidpat] (suspofden) {``suspectOfCOVID-19''};
		\node [FireBrick, left=of covid] (bot) {$\bot$};

		\path [arrow]
		(init) edge (s1)
		(init) edge[bend right = 70] (mary)
		(s1) edge (fevcough)
		(mary) edge (patient)
		(covid) edge[bend right = -45] (disease)
		(fevcough) edge (fever)
		(fevcough) edge (cough)
		(cough) edge (symptom)
		(fever) edge (symptom)
		(suspofden) edge (covidpat)
		;

		\path [arrow, NavyBlue]
		(mary) edge node [midway, fill=white, anchor=center] {hasSymptom} (s1)
		(patient) edge node [midway, fill=white, anchor=center] {hasSymptom} (symptom)
		(covidpat) edge node [midway, fill=white, anchor=center] {suspectOf} (covid)
		(symptom) edge[bend right = 45] node [midway, fill=white, anchor=center] {hasCause} (disease)
		;

		\path [arrow, dotted, FireBrick]
		(fevcough) edge node [midway, fill=white, anchor=center] {hasCause} (covid)
		(fever) edge node [midway, fill=white, anchor=center] {hasCause} (covid)
		;

		\path [arrow, dotted, FireBrick]
		(covid) edge (bot)
		;

	\end{tikzpicture}
	\caption{Graphical representation of the ontology in \autoref{exmp:gel-def}.}

	\label{fig:gelppGraph}
\end{figure}

\subsection{SAT decision}

\todo[inline]{Explain the algorithm of SAT decision and theorems}

- graph completion

- theorem: sat is path init - bot in completed graph

\section{MaxSAT for \texorpdfstring{$\gel$}{𝓖𝓔𝓛}}
\label{sec:maxgel}

Before focusing on the probabilistic extension for $\gel$, we need to define its MaxSAT problem, which will compose further the probabilistic reasoner.

The \emph{weighted partial maximum satisfiability problem for} $\gel$ ($\gel$-MaxSAT) can be defined as follows: given a potentially inconsistent weighted CBox $\cbox$, we want to find the maximal satisfiable subset of axioms; since it is partial, some axioms must be present in this subset. Usually, partiality can be modeled assigning infinite weights to the axioms that must not be excluded.

A \emph{weighted} CBox is a pair $\tuple{\cbox, w}$ where $\cbox$ is a CBox and $w: \cbox \to \Ratios \cup \{\infty\}$ is a weight function, which maps axioms in $\cbox$ to weights. The infinite weight is used to represent axioms that must not be excluded in the maximal satisfiable subset. Also, it is defined that RIs of the form $r \isa s$ and $r_1 \circ r_2 \isa s$ always have infinite weight.

Then, given a weighted CBox $\tuple{\cbox, w}$, a solution for the weighted partial $\gel$-MaxSAT problem is a set $\cbox_{max} \subseteq \cbox$ such that:
\begin{itemize}
	\item $\cbox_{max}$ is satisfiable; and
	\item $\cbox_{max} \models C \isa D$ if $w(C \isa D) = \infty$; and
	\item the sum of finite weights in $\cbox_{max}$ is maximal.
\end{itemize}

\section{Probabilistic \texorpdfstring{$\gel$}{𝓖𝓔𝓛}}
\label{sec:pgel}

Probability in $\gel$ is constructed from a \emph{probability function} $P$ \citep{Fin2020}. Consider a finite number of interpretation, $\I_1, \dots, \I_m$, we define the probability function $P: \{\I_1, \dots, \I_m\} \to \Ratios$, such that $P(\I_i) \geq 0$ and $\sum_{i = 1}^m P(\I_i) = 1$. We can also define the probability of an axiom $C \isa D$ as follows

\[
	P(C \isa D) = \sum_{\I_i \models C \isa D} P(\I_i).
\]

A \emph{probabilistic knowledge base} is a pair $\tuple{\CC, \PP}$, where $\CC$ is a CBox and $\PP$ is a PBox. A PBox is a set of $k$ linear constraints over $n$ axioms, of the form

\begin{equation}
\label{eq:pbox-restric}
	\sum_{j = 1}^n a_{ij} \cdot P(C_j \isa D_j) \leq b_i; \quad 1 \leq i \leq k.
\end{equation}

We can define the \emph{satisfiability problem} for this probabilistic KB (PGEL-SAT) as deciding if it is consistent or not. If it is consistent, the solution is a set of interpretations $\{\I_1, \dots, \I_m \}$ and a probability function $P: \{\I_1, \dots, \I_m \} \to \Ratios^+$ such that $\sum_{i = 1}^m P(\I_i) = 1$, $P(C \isa D) = 1$ for $C \isa D \in \CC$ (axioms in CBox are certain) and P verifies all linear constraints in $\PP$.
% 
\begin{example}
	Now we can model the uncertain situation stated in \autoref{exmp:real-example} using the probability knowledge base $\tuple{\CC_{exa}, \PP_{exa}}$, where $\CC_{exa}$ is the CBox from \autoref{exmp:gel-def} and $\PP_{exa}$ is given by
	\begin{align*}
		\PP_{exa} := \{ \quad P(Ax_2) - P(Ax_1) & = 0.2,           \\
		P(Ax_3)                                 & = 0.9  \quad \}.
	\end{align*}
	\todo[inline]{I don't think this is the right way to express a PBox}

	Then, we need a polynomial algorithm to find if this probabilistic KB is consistent.
\end{example}
% 
\subsection{Linear algebraic view}

The PGEL-SAT problem was also defined by \citet{Fin2020} in a linear algebraic view, which is useful to develop its polynomial reasoning algorithm. It was shown that a probabilistic KB $\tuple{\CC, \PP}$ is satisfiable iff the linear equation $C \cdot x = d$ has a solution $x \geq 0$, where 
% 
\begin{equation}
	\label{pbox-linear}
	C := \begin{bmatrix}
		 -I_n           & M_{n \times m} \\
		 A_{k \times n} & 0_{k \times m} \\
		 0'_n           & 1'_{m}
	\end{bmatrix}
	%
	\quad
	x := \begin{bmatrix}
		p_n \\
		\pi_m
	\end{bmatrix}
	%
	\quad
	d := \begin{bmatrix}
		0_n \\
		b_k \\
		1
	\end{bmatrix}
\end{equation}

and
\begin{itemize}
	\item $A_{k \times n}$ is a $k \times n$ matrix whose elements $a_{ij}$ are given by \autoref{eq:pbox-restric};
	\item $b_k$ is a $k$ vector whose elements $b_i$ are also given by \autoref{eq:pbox-restric};
	\item $M_{n \times m}$ is a $n \times m$ matrix given by the following steps:
	
	Consider an interpretation $\I$ model of $\CC$, its corresponding vector in $\PP$ is  a $\{0, 1\}$-vector $y$ such that $y_i = 1$ iff $\I \models C_i \isa D_i$, for $1 \leq i \leq n$.

	Then, given a set of interpretations $\I_1, \dots, \I_m$, we define $M_{n \times m}$ a matrix whose column $M^j$ is $\I_j$'s corresponding vector in $\PP$; 
	\item $I_n$ is the $n$-dimensional identity matrix;
	\item $0_n$ is a column $0$-vector of size $n$ (similarly for $1_n$);
	\item $0'_n$ is the previous vector's transpose;
	\item $0_{k \times m}$ is a $0$-matrix of shape $k \times m$; 
	\item $p_n$ is a vector of size $n$ which corresponds to the probability of axioms occurring in \autoref{eq:pbox-restric}. 
	\item $\pi_m$ is a vector of size $m$ which corresponds to the probability distribution over interpretations $\I_1, \dots, \I_m$.
\end{itemize}

\todo[inline]{Maybe define ill-formed and well-formed columns; Definition of cost vector; Definition of the linear system solution by the minimation problem}

% definition

% column generation algorithm

% example